---
title: "Springboard Foundations of Data Science - Milestone Report"
author: "Susan Joseph"
date: "November 5, 2017"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE)
library(Zelig)
library(effects)
library(dplyr)
library(ggplot2)
```




##Introduction
###An anonymous bank based in Spain has published 1.5 years of customer behavior data  that consists of customer personal details (Customer ###Personal Details.csv) and the products of the bank they have subscribed to (Customer Products.csv). The different products that the ###customers subscribed to are Current Account, Particular Account, Direct Debit, E-account, Payroll Account, Taxes , Credit Cards, ###Pensions, Securities, Funds and Long Term Deposits. The major Customer attributes that are of interest are the Age, household income, the ###channel used by the customer to subscribe to a product and Segmentation 01 - VIP, 02 - Individuals 03 - college graduated. The goal of ###the project is to determine if the Customer attributes mentioned have any correlation to the different bank products so that the ###marketing department of the Bank could target new customers based on the analysis.


##Approach
###The capstone project is approached by first identifying the problem statement. Next, the most important csv files are identified. Data ###wrangling is done to tidy the data.  After this imputations , exploratory data analysis and some basic statistical correlations are ###performed. Finally, regressions and machine learning algorithms are used on the dataset.


##Problem Statement
###The goal of the project is to determine if there are any correlations between the  Customer attributes and the correlation to the ###different bank products so that the marketing department of the Bank could target new customers.


##Why and for whom is this problem statement important
###The problem is important to the Bank in Spain to find out what customer attributes such as Age, Income and Segment would be responsible ###in influencing the subscription ###of the bank product by the customer.

##Data
###The Bank has provided two .CSV files Customer Personal Details.csv and Customer Products.csv. The data starts at 2015-01-28 and has ###monthly records of products a customer has, such as "credit card", "savings account", etc. These products are the columns named: ###ind_(xyz)_ult1 etc. There are 1048576 rows and 24 observations in the Customer Peronal Details.csv file  and 1048576 rows and 24 ###observations in the Customer Products.csv file.


```{r echo=FALSE, warning=FALSE , include=FALSE}


customer <- read.csv('Customer Personal Details.csv', 
                     stringsAsFactors = FALSE, na.strings = "")

product <- read.csv('Customer Products.csv', 
                     stringsAsFactors = FALSE, na.strings = "")

```
##Structure of the Customer and Product files

```{r echo=FALSE, warning=FALSE}

str(customer)

str(product)
```
##Converting the variables to the correct data types

```{r echo=FALSE, warning=FALSE}
customer$fecha_dato<-as.Date(customer$fecha_dato)
customer$ind_empleado<-as.factor(customer$ind_empleado)
customer$sexo<- as.factor(customer$sexo)
customer$age <- as.numeric(customer$age)
customer$ind_nuevo<- as.factor(customer$ind_nuevo)
customer$indfall<- as.factor(customer$indfall)
customer$renta<-as.numeric(customer$renta)
customer$ind_actividad_cliente<- as.factor(customer$ind_actividad_cliente)
customer$segmento<-as.factor(customer$segmento)
product$ind_nomina_ult1<-as.numeric(product$ind_nomina_ult1)
product$ind_nom_pens_ult1<-as.numeric(product$ind_nom_pens_ult1)

```
##Data Wrangling

###There are 17 rows for one Customer where every row is a month of data of a customer. The below ouput displays the number of rows for each customer
```{r echo=FALSE, warning=FALSE}
group_by(customer, ncodpers) %>%
  summarise(n = n()) %>%
  group_by(n)
```

###Summarizing the rows so there is only one row per customer and eliminating columns that do not add value to the analysis. There are ###78,359 customers after the summarization

```{r echo=FALSE, warning=FALSE}

cust_data <- group_by(customer, ncodpers) %>%
  summarise(
    ind_empleado = last(ind_empleado),
    pais_residencia = last(pais_residencia),
    sexo = last(sexo), 
    age = last(age),
    fecha_alta = last(fecha_alta), 
    antiguedad = last(antiguedad), 
    indrel = last(indrel),
    indrel_1mes = last(indrel_1mes),
    tiprel_1mes = last(tiprel_1mes),
    indresi = last(indresi),
    indext = last(indext), 
    canal_entrada = first(canal_entrada),
    ind_actividad_cliente = last(ind_actividad_cliente),
    renta = mean(renta, na.rm=TRUE), 
    segmento = last(segmento)
  )

```

### Replacing the column canal_entrada with 'Other' values where count of observations is less than 200

```{r echo=FALSE, warning=FALSE}
canal_entrada_keep <- group_by(cust_data, canal_entrada) %>%
  summarise(n = n()) %>% filter(n > 200)

canal_entrada_keep <- canal_entrada_keep$canal_entrada

cust_data$canal_entrada_clean <- ifelse(cust_data$canal_entrada %in% canal_entrada_keep, cust_data$canal_entrada, 'OTHER')
```

### Hence all the  categories have more than 200 elements
```{r echo=FALSE, warning=FALSE}
table(cust_data$canal_entrada_clean)

cust_data$canal_entrada_clean <- as.factor(cust_data$canal_entrada_clean)

cust_data <-cust_data %>% select (-canal_entrada)
```

### Finding out the number of NA's by column 
```{r echo=FALSE, warning=FALSE}
apply(cust_data, MARGIN = 2, function(x) sum(is.na(x)))
```

### Impute NA's for each column

```{r echo=FALSE, warning=FALSE}
table(cust_data$sexo)
```
#### sexo: Since most of the values are "V" then we will suppose that the 15 missing are also from that category

```{r echo=FALSE, warning=FALSE}
cust_data$sexo[is.na(cust_data$sexo)] <- "V"
```
#### renta: This one is complicated because about 20% of the values are missing, however for the sake of simplicity we will do 
#### the easiest solution and take the median of the values

```{r echo=FALSE, warning=FALSE,include=FALSE}
table(cust_data$renta)
cust_data$renta[is.na(cust_data$renta)] <- median(cust_data$renta, na.rm = TRUE)
```
#### segmento: Since most of the values are from the 02-PARTICULARES segment we will suppose that 1072 missing entries are also from that #### segment
```{r echo=FALSE, warning=FALSE}
table(cust_data$segmento)
cust_data$segmento[is.na(cust_data$segmento)] <- "02 - PARTICULARES"
```

#### canal_entrada_clean: Since most of the values are from the KHE Channel we will assume that the 1005 missing entries are also from the #### KHE channel

```{r echo=FALSE, warning=FALSE}
table(cust_data$canal_entrada_clean)
cust_data$canal_entrada_clean[is.na(cust_data$canal_entrada_clean)]<- "KHE"
```

#### ind_empleado: Since most of the Employee Index values are of type 'N'(Not employee)  we will assume that the 514 missing entries are #### also of type 'N'
```{r echo=FALSE, warning=FALSE}
table(cust_data$ind_empleado)
cust_data$ind_empleado[is.na(cust_data$ind_empleado)] <- "N"
```

#### pais_residencia: Since most of the Employees are from Spain (ES) we will assume that the 514 missing entries are also from the country #### Spain
```{r echo=FALSE, warning=FALSE}
table(cust_data$pais_residencia)
cust_data$pais_residencia[is.na(cust_data$pais_residencia)] <- "ES"
```
####indrel_lmes: Since most of the Customer type is 1.0 we will assume that the 735 missing entries are also of the same Customer type
```{r echo=FALSE, warning=FALSE}
table(cust_data$indrel_1mes)
cust_data$indrel_1mes[is.na(cust_data$indrel_1mes)] <- "1.0"
```

####tiprel_lmes: Since most of the Customer relation is  of type 'A'(Active) we will assume that the 735 missing entries are also of the ####same Customer relation type
```{r echo=FALSE, warning=FALSE}
table(cust_data$tiprel_1mes)
cust_data$tiprel_1mes[is.na(cust_data$tiprel_1mes)] <- "A"
```

####age: Since most of the Customers are of age 23 we will assume that the 514 missing entries are also of the same age
```{r echo=FALSE, warning=FALSE}
table(cust_data$age)
cust_data$age[is.na(cust_data$age)] <- 23
```


####indresi: Since most of the Customers are residents of Spain with a resident index of 'S' we will assume that the 514 missing entries are ####also residents 
```{r echo=FALSE, warning=FALSE}
table(cust_data$indresi)
cust_data$indresi[is.na(cust_data$indresi)] <- "S"
```

####fecha_alta: This one was a tough one but since there were 15 rows for 1995-10-27 we will assume that the 514 missing entries are also in ####that year and month when the customer became the first holder of a contract in the bank.
```{r echo=FALSE, warning=FALSE, include=FALSE}
table(cust_data$fecha_alta)
cust_data$fecha_alta[is.na(cust_data$fecha_alta)] <- "1995-10-27"
```


####indext: Since most of the customers are born in Spain and is the same as the bank country the foreigner index is 'N' and the missing 514 ####entries are also 'N'
```{r echo=FALSE, warning=FALSE}
table(cust_data$indext)
cust_data$indext[is.na(cust_data$indext)] <- "N"
```

### Finally all the columns are zeroes as all the NA's in all the columns were imputed and replaced
```{r echo=FALSE, warning=FALSE}
apply(cust_data, MARGIN = 2, function(x) sum(is.na(x)))
```

##Descriptive and Exploratory Analysis

####The number of males are more than females at 42,403
```{r echo=FALSE, warning=FALSE}
table(cust_data$sexo)
```

#### The percentage of males are more than females at 54.12%
```{r echo=FALSE, warning=FALSE}
100*round(table(cust_data$sexo)/sum(table(cust_data$sexo)),5)
```


####Plotting the sex in a Histogram shows that the count of males are more than females
```{r echo=FALSE, warning=FALSE}
ggplot(cust_data, aes(x=sexo)) +
  geom_bar()
```

#### Age: The minumum age of the customer is 2 and maximum is 117. The mean or average age of the customer is around 38 years and the
#### standard deviation is only 16.98 which means they are closer to the mean age.
```{r echo=FALSE, warning=FALSE}
cust_data$age <- as.numeric(cust_data$age)
summary(cust_data$age)
mean(cust_data$age)
sd(cust_data$age)
median(cust_data$age)
min(cust_data$age)
max(cust_data$age)
```

#### Plotting the Customer Sex against Age in a Box plot shows that 50% of the females are in the age range 25 to 45 and males are between #### 27 and 55.  The median age for females is around 30 and that of males is around 40
```{r echo=FALSE, warning=FALSE}
ggplot(cust_data, aes(x=sexo, y=age)) +
  geom_boxplot()
```


####Channel: Customer use KHE Channel the most with a total of 37,322 and  account for 47.63% of the total
```{r echo=FALSE, warning=FALSE}
table(cust_data$canal_entrada_clean)
100*round(table(cust_data$canal_entrada_clean)/sum(table(cust_data$canal_entrada_clean)),5)
```

#### Gross income of household: The mean income is 121,388 . The minimum salary of the Customers is 5341 and the maximum salary is
#### 15,711,716
```{r echo=FALSE, warning=FALSE}
summary(cust_data$renta)
mean(cust_data$renta)
sd(cust_data$renta)
median(cust_data$renta)
min(cust_data$renta)
max(cust_data$renta)
```

#### Customer Segmentation: The highest segment of customers are 02 - PARTICULARES(Individuals) with 41047 and accounting for 52.38% as 
#### shown by the bar graph
```{r echo=FALSE, warning=FALSE}
table(cust_data$segmento)
100*round(table(cust_data$segmento)/sum(table(cust_data$segmento)),5)
ggplot(cust_data, aes(x=segmento )) +
  geom_bar()
```




